### 2015-05-03 Rmarkdown/comethylation_review.Rmd
Edits to front matter and table of contents

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star_fourier.pdf}
\caption[Fourier transformation of incorrectly defined $\beta$-values]{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \textit{FF} sample. Fourier transformation computed and plotted using the \texttt{spectrum} function in R. The period is defined as 1 / frequency. The maximum frequency is approximately $0.33$, which corresponds to a period of approximately $3$ bp.}
\label{fig:FF_chr21_acf_beta_star_fourier}
\end{figure}

@@

\begin{table}[h]
\centering
\caption[Notions of `subject' and `population' in methylation linkage]{Notions of `subject' and `population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}. NB `person' could be any organism, e.g, `mouse', `dog', etc.}
\label{tab:linkage}
\begin{tabular}{@{}lll@{}}
\toprule
                    & Subject               & Population \\ \midrule
Genetic linkage     & Person                   & Population \\
Methylation linkage & (Haploid) genome of cell & Person     \\ \bottomrule
\end{tabular}
\end{table}

>>>>

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star_fourier.pdf}
\caption[Fourier analysis of incorrectly defined $\beta$-values]{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \textit{FF} sample. Fourier transformation computed and plotted using the \texttt{spectrum} function in R. The period is defined as 1 / frequency. The maximum frequency is approximately $0.33$, which corresponds to a period of approximately $3$ bp.}
\label{fig:FF_chr21_acf_beta_star_fourier}
\end{figure}

@@

\begin{table}[h]
\centering
\caption[Notions of `subject' and `population' in methylation linkage analyses]{Notions of `subject' and `population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}. NB `person' could be any organism, e.g, `mouse', `dog', etc.}
\label{tab:linkage}
\begin{tabular}{@{}lll@{}}
\toprule
                    & Subject               & Population \\ \midrule
Genetic linkage     & Person                   & Population \\
Methylation linkage & (Haploid) genome of cell & Person     \\ \bottomrule
\end{tabular}
\end{table}

### 2015-05-03 Rmarkdown/methsim.Rmd
Edits to front matter and table of contents

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_by_pm_boxplots.pdf}
\caption[\emph{EPISCOPE}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{EPISCOPE} dataset.}
\label{fig:EPISCOPE_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Lister_beta_by_pm_boxplots.pdf}
\caption[\emph{Lister}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Lister} dataset.}
\label{fig:Lister_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
\caption[\emph{Seisenberger}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Seisenberger} dataset.}
\label{fig:Seisenberger_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_beta_by_pm_boxplots.pdf}
\caption[\emph{Ziller}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Ziller} dataset.}
\label{fig:Ziller_beta_by_pm_boxplots}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly.pdf}
\caption[\texttt{methsim} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for \texttt{methsim} simulated data and the \emph{ADS} input sample. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10 \times$ sequencing coverage are included.}
\label{fig:methsim_plotBetaFreqPoly}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly_CGI.pdf}
\caption[\texttt{methsim} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10 \times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:methsim_plotBetaFreqPoly_CGI}
\end{figure}

>>>>

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_by_pm_boxplots.pdf}
\caption[\emph{EPISCOPE}: $\beta$-values by region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{EPISCOPE} dataset.}
\label{fig:EPISCOPE_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Lister_beta_by_pm_boxplots.pdf}
\caption[\emph{Lister}: $\beta$-values by region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Lister} dataset.}
\label{fig:Lister_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
\caption[\emph{Seisenberger}: $\beta$-values by region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Seisenberger} dataset.}
\label{fig:Seisenberger_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_beta_by_pm_boxplots.pdf}
\caption[\emph{Ziller}: $\beta$-values by region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Ziller} dataset.}
\label{fig:Ziller_beta_by_pm_boxplots}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly.pdf}
\caption[\texttt{methsim}: $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for \texttt{methsim} simulated data and the \emph{ADS} input sample. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10 \times$ sequencing coverage are included.}
\label{fig:methsim_plotBetaFreqPoly}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly_CGI.pdf}
\caption[\texttt{methsim}: $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10 \times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:methsim_plotBetaFreqPoly_CGI}
\end{figure}

### 2015-05-03 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Edits to front matter and table of contents

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption[\texttt{Bis-SNP} method schematic]{\texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveals the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

>>>>

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption[Schematic of \texttt{Bis-SNP} method]{\texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveals the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

### 2015-05-03 Rmarkdown/wgbs_statistical_framework.Rmd
Edits to front matter and table of contents

\subsection{Some complications for a single sample}\label{sec:one_sample_complications}

@@

\subsection{Some complications with multiple samples}\label{sec{multi_sample_complications}}

>>>>

\subsection{Some complications}\label{sec:one_sample_complications}

@@

\subsection{Some complications}\label{sec{multi_sample_complications}}

### 2015-05-03 latex/phd_thesis.tex
Edits to front matter and table of contents

\title{\huge \bf The statistical analysis of high-throughput assays
for studying DNA methylation.\\}

@@

\date{The Walter and Eliza Hall Institute of Medical Research}

@@

This thesis examines the statistical and computational challenges raised by high-throughput assays of DNA methylation, particularly the current gold standard assay of whole-genome bisulfite-sequencing. Using whole-genome bisulfite-sequencing, we can now measure DNA methylation at individual nucleotides across entire genomes. These experiments produce vast amounts of data that require new methods and tools to analyse.

@@

The second half of the thesis focuses on \emph{co-methylation}, the spatial dependence of DNA methylation. We demonstrate that previous analyses of co-methylation have been limited by inadequate data and deficiencies in the statistical methods applied. This motivates a study of co-methylation from 40 whole-genome bisulfite-sequencing samples. These 40 samples represent a diverse range of tissues, from embryonic and induced pluripotent stem cells, through to somatic cells and tumours. Making use of software developed in the first half of the thesis, we explore different measures of co-methylation and relate these to one another. We identify genomic features that influence co-methylation and how it varies between different tissues.

@@

I would like to thank my PhD supervisors, Terry Speed and  Peter Hall. Peter, thank you for supporting my decision to radically change tack very early on in my PhD from a topic in theoretical statistics to a heavily computational and interdisciplinary problem. And Terry, thank you for agreeing to take me on as a student after this change of heart. I have learnt much from working with you and greatly appreciate your generosity and guidance over the course of my PhD. Thanks also to Gordon Smyth, the chair of my PhD committee, for always taking the time to help me.

>>>>

\title{\huge \bf The statistical analysis of high-throughput assays
for studying DNA methylation\\}

@@

\date{Bioinformatics Division\\
The Walter and Eliza Hall Institute of Medical Research}

@@

This thesis examines the statistical and computational challenges raised by high-throughput assays of DNA methylation, particularly the current gold standard assay of whole-genome bisulfite-sequencing. Using whole-genome bisulfite-sequencing, we can now measure DNA methylation at individual nucleotides across entire genomes. These experiments produce vast amounts of data that require new methods and software to analyse.

@@

The second half of the thesis focuses on \emph{co-methylation}, the spatial dependence of DNA methylation along the genome. We demonstrate that previous analyses of co-methylation have been limited by inadequate data and deficiencies in the applied statistical methods. This motivates a study of co-methylation from 40 whole-genome bisulfite-sequencing samples. These 40 samples represent a diverse range of tissues, from embryonic and induced pluripotent stem cells, through to somatic cells and tumours. Making use of software developed in the first half of the thesis, we explore different measures of co-methylation and relate these to one another. We identify genomic features that influence co-methylation and how it varies between different tissues.

@@

I would like to thank my PhD supervisors, Terry Speed and  Peter Hall. Peter, thank you for supporting my decision to radically change tack very early on in my PhD from a topic in theoretical statistics to a heavily computational and interdisciplinary problem. Terry, thank you for agreeing to take me on as a student after this change of heart. I have learnt much from working with you and greatly appreciate your generosity and guidance over the course of my PhD. Thanks also to Gordon Smyth, the chair of my PhD committee, for always taking the time to help me.
